Do investors in clean energy ETFs herd? The role of climate risks

### 2025-06-25 alt_energy_herding.tex
added from comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={blue!100!black},
%     citecolor={blue!100!black},
%     urlcolor={blue!100!black}
% }
% \usepackage{apacite}
% \usepackage[round]{natbib} 
% \usepackage{graphicx}
% \usepackage{float}
% \usepackage{caption}
% \usepackage[toc,page]{appendix}
% \usepackage{booktabs,caption}
% \usepackage[flushleft]{threeparttable}
% \usepackage{tabularx}
\usepackage{sfmath}
\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{amsmath}
% \usepackage{xcolor}
% \usepackage{scrextend}
% \deffootnote[2em]{2em}{1em}{\textsuperscript{\thefootnotemark}\,}
% \newcolumntype{Y}{>{\centering\arraybackslash}X}
% \usepackage[a4paper]{geometry}
% \usepackage{caption}
% \usepackage[bottom,flushmargin,hang,multiple]{footmisc}
% \usepackage{pdflscape}
% \usepackage[paper=portrait,pagesize]{typearea}
% \usepackage{rotating}
% \usepackage{fullpage}
% \usepackage{tabu}
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{rotating}
\newcommand{\bsideways}{\begin{sidewaystable}[htbp]}
\newcommand{\esideways}{\end{sidewaystable}}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[round]{natbib}
\bibliographystyle{apalike}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={blue!100!black},
  filecolor={Maroon},
  citecolor={blue!100!black},
  urlcolor={blue!100!black},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\title{Do investors in clean energy ETFs herd?}

@@

\noindent\textbf{Keywords}: Herding Behaviour, Climate Change, Energy
\\
\textbf{JEL Codes}: G14, Q54, P18
\newpage

\setstretch{1.5}
\section{Introduction and literature
review}\label{introduction-and-literature-review}

Climate sustainability is a major concern for financial markets and
investors. The growing interest in climate sustainability is primarily
driven by the increasing materialisation of climate risks, and the
actions of governments, institutions and organizations towards a
sustainable future \citep{giglio2021a}. To ensure returns, investors
increasingly seek to hedge against climate risks by investing in green
financial products. Although the evidence is mixed, indications that
returns from green financial products are comparable to traditional
financial \citetext{\citealp[see amongst
others,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.

In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
these climate risks, a general change in investor attitudes can drive
the inclusion of green assets in their portfolio can lead to systemic
risk.\footnote{These pressures notwithstanding the possible contribution
  that financial markets can play in mitigating and reducing the
  negative effects of climate change \citep{giglio2021a}}

Specifically, climate change presents risks to investor portfolios
through two primary sources - physical and transition risks. Physical
risks or direct impact refer to extreme climate events such as floods
and droughts, which impact business operations and infrastructure; and
transition risks are the policy, technological, and other costs that
societies bear to achieve low carbon economies
\citetext{\citealp{nguyen2025}; \citealp[and][ amongst
others]{giglio2021a}}. Investors, therefore, recognise these risks and
seek to mitigate them, as they seek return-enhancing green financial
products.

Exchange-traded funds (ETFs) are a key feature of green financial
products. ETFs are a type of security that involves a collection of
securities that often tracks an underlying index. However, they can
invest in various industry sectors or strategies. In addition,
environment, social, and governance (ESG) ETFs serve as a market
discovery tool for investors to identify and invest in environmentally
friendly companies \citep{briere2023}. Among the ESG ETFs, the Clean
Energy (CE) ETFs have been the best-performing ones in 2022, followed by
the Cybersecurity and Artificial Intelligence (AI) ETFs
\citep{decclesia2024}. The clean energy transition represents one of the
largest multi-decade secular growth opportunities. After the inclusion
of Green energy financing in the list of United Nations Sustainability
Goals (SDGs) as SDG 7, the role, importance, and visibility of green
financial products have escalated enormously \citep{naqvi2022}. That is,
the growth of green assets under management is likely to continue.
Lastly, the limited availability of data on ESG complying investment
tools \citep{avramov2022, nguyen2025} justifies the use of green (CE)
ETFs as best candidates of green assets.

However, given how recent the inclusion of climate sustainability in
investment decisions is, it is not clear what the actual impact will be
in the long run. In this study specifically, we ask whether the rapid
adoption of CE ETFs could be driven by market fads, or is a fundamental
change in investor behaviour. Investors, for example, can believe that
peers have information about climate risks that they do not, investors
may herd to avoid losses compared to peers, or investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Therefore,
market volatility and crisis, financial performance, and investor
sentiment can drive herding in ESG markets.

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behavior can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia,][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep{lakonishok1992}, and studies that use market returns data and
investigate herding towards the market consensus
\citep{chang2000, galariotis2015}. Our paper falls within the latter
category and tests for herding towards the market consensus for clean
energy US ETFs.

Herding behaviour in ESG markets is not without precedent. Amongst
others, \citet{loang2023} found that compliance with SDG goals can
introduce bias in investor sentiment, which leads to herding behaviour.
Using a Twitter (or X) uncertainty index, \citet{koutmos2024} found
evidence of herding in US-based ESG index fund investors.
\citet{przychodzen2016} found herding behaviour amongst fund managers
who incorporated ESG strategies in their portfolios. Lastly,
\citet{rubbaniy2021} highlight evidence of herding in the MSCI US ESG
Leader Index during extreme periods (bear and bull periods).

However, to the best of our knowledge, no studies exist that focus on
herding behaviour in CE ESGs. This study, therefore, aims to extend the
broader literature on herding behaviour in commodity and energy markets.
Several studies in this area were conducted. For example,
\citet{demirer2013a} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} showed herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} found significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} showed that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} also found no herding behaviour in
international commodity markets. Overall, the literature in this area is
mixed, which indicates scope for further study.

In addition, our study extends the results of
\citet{dragomirescu-gaina2021} who examined herding behavior of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They employed the
energy equities included in the S\&P 500 and concluded that herding
among investors in the US energy market sector is sensitive to green
volatility shocks

Therefore, this study investigates herding behaviour in alternative
energy ETFs in the US between 2016 and 2024. We then demonstrate how
climate-related uncertainty can drive herding behaviour in these
markets. Methodologically, we follow the standard herding tests by
\citet{christie1995} and \citet{chang2000}. The traditional approach was
supplemented by quantile regressions \citep{koenker1978} in order to
capture the time-varying aspects of herding. Lastly, we extend recent
approaches by \citet{bua2024} and others, which seek to establish a link
between climate uncertainty and herding behaviour.

@@

\includegraphics{alt_energy_herding_files/figure-pdf/fig-csad-1.pdf}

@@

Following the relevant literature
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}, we
compute dispersion of the \(i^{th}\) ETF from the market return. This is
known as the Cross Sectional Absolute Deviation (\(CSAD_t\)) measure.
Empirically the \(CSAD_t\) is defined in the following manner:

@@

\includegraphics{alt_energy_herding_files/figure-pdf/fig-rolling_window-1.pdf}

@@

We observe several periods of herding behaviour as reflected in the
troughs in Figure~\ref{fig-rolling_window}. The most prominent cases of
herding occur between March and May of 2020 followed by several
instances of herding in the period that extends from March through April
of 2017 (when \(\gamma_2\) averaged -45.5) and the period of
February-March of 2023 (when \(\gamma_2\) averaged -7.68).We observe
significant herding in the onset of the COVID-19 crisis 2020, which was
short lived as market adapted. The rest of the Covid period was in the
main characterised by anti-herding behaviour. \citet{ghorbel2023} and
\citet{dhall2020} found similar results. Lastly, similar to
\citet{mohamad2024} we observe no herding behaviour related to the
Russia-Ukraine war. \citet{mohamad2024} highlights the importance of
context in understanding the link between herding and market events. On
the other side, we derive significant moments of anti-herding behaviour
in the clean energy ETFs by observing the spikes in
Figure~\ref{fig-rolling_window}. Cross sectional dispersion appears to
increase with respect to market-wide returns which is a sign of
anti-herding behaviour on behalf of investors around December of 2016
and later during September of 2022.

\subsection{Climate-related uncertainty and herding
behaviour}\label{climate-related-uncertainty-and-herding-behaviour}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[round]{natbib}
\bibliographystyle{apalike}

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={blue!100!black},
%     citecolor={blue!100!black},
%     urlcolor={blue!100!black}
% }
% \usepackage{apacite}
% \usepackage[round]{natbib} 
% \usepackage{graphicx}
% \usepackage{float}
% \usepackage{caption}
% \usepackage[toc,page]{appendix}
% \usepackage{booktabs,caption}
% \usepackage[flushleft]{threeparttable}
% \usepackage{tabularx}
\usepackage{sfmath}
\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{amsmath}
% \usepackage{xcolor}
% \usepackage{scrextend}
% \deffootnote[2em]{2em}{1em}{\textsuperscript{\thefootnotemark}\,}
% \newcolumntype{Y}{>{\centering\arraybackslash}X}
% \usepackage[a4paper]{geometry}
% \usepackage{caption}
% \usepackage[bottom,flushmargin,hang,multiple]{footmisc}
% \usepackage{pdflscape}
% \usepackage[paper=portrait,pagesize]{typearea}
% \usepackage{rotating}
% \usepackage{fullpage}
% \usepackage{tabu}
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{rotating}
\newcommand{\bsideways}{\begin{sidewaystable}[htbp]}
\newcommand{\esideways}{\end{sidewaystable}}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  colorlinks=true,
  linkcolor={blue!100!black},
  filecolor={Maroon},
  citecolor={blue!100!black},
  urlcolor={blue!100!black},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}
\begin{document}

\title{Do investors in clean energy ETFs herd? The role of climate risks}

@@

\noindent\textbf{Keywords}: Herding Behaviour, Climate Change, Clean Energy
\\
\textbf{JEL Codes}: G14, Q54, P18
\newpage

\setstretch{1.5}
\section{Introduction}\label{introduction}

Climate sustainability is a major concern for financial markets and
investors. The growing interest in climate sustainability is primarily
driven by the increasing materialisation of climate risks, and the
actions of governments, institutions and organizations towards a
sustainable future \citep{giglio2021a}. To safeguard returns, investors
increasingly seek to hedge against climate risks by investing in green
financial products. Although the evidence from the existing literature
is mixed, returns from green financial products are somewhat comparable
to traditional financial products \citetext{\citealp[see amongst
others,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.
In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
climate risks, a general change in investor attitudes can drive the
inclusion of green assets in their portfolio can lead to systemic
risk.\footnote{These pressures notwithstanding the possible contribution
  that financial markets can play in mitigating and reducing the
  negative effects of climate change \citep{giglio2021a}} Specifically,
climate change presents risks to the global financial system, and
ultimately to investment portfolios, through two primary sources -
physical and transition risks. Physical risks or direct impact refer to
extreme climate events such as floods and droughts, which impact
business operations and infrastructure; and transition risks are the
policy, technological, and other costs that societies bear to achieve
low carbon economies \citetext{\citealp{nguyen2025}; \citealp[and][
amongst others]{giglio2021a}}. Investors, therefore, recognise these
risks and seek to mitigate them, as they seek return-enhancing green
financial products.

Energy market has been in the epicentre of retail investors and
portfolio managers at least over the last two decades. Official
estimates claim that global energy investment is expected to surpass USD
3 trillion for the first time in 2024, with USD 2 trillion going to
clean energy technologies and infrastructure \citep{iea2024}. There is a
general consensus in the literature that the most important event that
has shaped the current landscape in the energy sector is the emergence
of the energy portfolio with the rise of green energy as an alternative
energy resource and an asset class. Price volatility of fossil fuels on
the one hand and decarbonisation-related goals dictated by international
agreements on the other hand are heavily responsible for the growing
interest in green energy market. Responding to the increasing awareness
about the environment and the need for more green capital, financial
markets have also attracted a large portion of retail and institutional
investors' fund flows \citep{rizvi2022green, naqvi2021there}.

Needless to mention that the rise of energy popularity as an alternative
investment class has also been fueled by the fictionalization of
commodities markets. Commodities have become an effective investment
tool for investment community as it allows investors to participate in
the market developments across various physical products such as
agricultural products, silver, gold, natural gas, oil etc and more
recently clean energy products without being obliged to own the product
directly \citep{gagnon2020they}. Several academic studies have
highlighted substantial diversification benefits of commodities when
added to a portfolio of stocks and bonds \citep[see inter
alia,][]{naqvi2022}.

Green energy sectors like solar, wind, geothermal, biofuels, biomass,
wave, hydro and tidal energies are considered by investors, authorities
and policy-makers as the substitutes for fossil-fuels energy that
increase CO2 emissions \citep{shahzad2020clean}. Moreover, as
\citet{naqvi2022} point out that ``the alternative energy sector is
expected to grow faster over the next decades due to the depletion of
fossil fuel reserves and the technological advance that has made green
energy feasible''.

Although green energy investment has witnessed an exponential growth
over the years, there is a lot of improvement that can be done in terms
of investors' participation, both retail and institutional. To achieve
the goal of larger capital commitment to green energy projects it is
important to gain investors' confidence in financial products that are
linked to green energy such as Exchange-traded funds(ETFs).

ETFs are a key feature of green financial products. They are a
collection of investments that often tracks an underlying performance of
an asset or index. In the universe of environment, social, and
governance (ESG) investments, clean ETFs serve as a tool for
environmentally conscious investors to identify and invest in
environmentally friendly companies \citep{briere2023} engaged in the
transition towards a cleaner production and a low-carbon economy. Among
the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing
one in 2022 \citep{decclesia2024}. This is not surprising given that the
clean energy transition represents one of the largest multi-decade
secular growth opportunities. After the inclusion of Green energy
financing in the list of United Nations Sustainability Goals (SDGs) as
SDG 7, the role, importance, and visibility of green financial products
have escalated enormously \citep{naqvi2022}. That is, the growth of
green assets under management is likely to continue. Furthermore, the
limited availability of data on ESG complying investment tools
\citep{avramov2022, nguyen2025} justifies the use of energy ETFs as best
candidates of green assets. However, given that the inclusion of climate
sustainability in investment decisions is a recent phenomenon, it is not
clear what the actual impact will be in the long-run.

This study investigates herding behaviour in alternative (clean) energy
ETFs in the US between May 1 2016 and June 19 2024, showing evidence of
significant herding that is asymmetric and time-varying. We then study
whether climate-related uncertainty can affect the herding behaviour in
the US clean energy ETFs market. Methodologically, we follow the
standard herding tests by \citet{christie1995} and \citet{chang2000}.
Notably, we supplement the traditional approach with quantile
regressions \citep{koenker1978} in order to capture how herding differs
across various quantiles of the returns dispersion and in up and down
markets. Lastly, to establish a link between climate risk and herding
behaviour, we differentiating between transitional and physical risks
\citep{bua2024} and provide valuable insights on their potential impact
on the likelihood of herding in clean energy ETFs.

In this study specifically, we ask whether the rapid adoption of clean
energy ETFs could be driven by market fads, or is a fundamental change
in investor behaviour. Investors, for example, can believe that their
peers have more valuable information about climate risks, making them to
herd to avoid losses compared to peers; alternatively, investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Investors in
clean energy stocks can experience large losses if their betting on
clean energy stocks go wrong, which represents a major concern in their
decision-making process. This incites those investors to disregard their
own information and follow the market consensus, leading to significant
herding behaviour in the clean energy market. \citet{devenow1996}
indicate that following the market consensus induces some kind of
security among less informed traders. This could be relevant to our
analysis because ETFs represent a popular investment instrument for
individual and retail investors who are not necessarily well informed
about the risk and prospects of investments in clean energy firms and
whether their contribution to the world transition to cleaner production
should be financially lucrative. On a related front, the
self-reinforcing nature of confidence in the tyranny of the majority, as
indicated by \citet{teraji2003} could also be pertinent.

While herding behaviour has been examined in ESG markets, it remains
understudied in clean energy assets, notably, clean energy
ETFs.\footnote{Amongst others, \citet{loang2023} shows that compliance
  with SDG goals can introduce bias in investor sentiment, which leads
  to herding behaviour. Using a Twitter (or X) uncertainty index,
  \citet{koutmos2024} finds evidence of herding in US-based ESG index
  fund investors. \citet{przychodzen2016} found indicate evidence of
  herding behaviour amongst fund managers who incorporated ESG
  strategies in their portfolios. Lastly, \citet{rubbaniy2021} highlight
  evidence of herding in the MSCI US ESG Leader Index during extreme
  (bear and bull periods) periods.} Other studies on herding have been
conducted in commodity and fossil energy markets. For example,
\citet{demirer2013} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} shows herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} find significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} show that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} find no herding behaviour in
international commodity markets. Notably, our study extends the results
of \citet{dragomirescu-gaina2021} who have examined herding behavior of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They study the energy
equities included in the S\&P 500 and concluded that herding among
investors in the US energy market sector is sensitive to green
volatility shocks

Our analysis shows that herding is significant and is present in both
down and up markets, with a stronger effect in the down market,
suggesting an asymmetry. Herding is also found to be time-varying.
Notably, an additional analysis reveals that the transition climate
risk, particularly its high level, reduces the probability of herding in
clean energy ETFs, whereas physical climate risk does not exert any
significant impact on the probability of herding

@@

\pandocbounded{\includegraphics[keepaspectratio]{alt_energy_herding_files/figure-pdf/fig-csad-1.pdf}}

@@

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behavior can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep[e.g.][]{lakonishok1992}, and studies that use market returns data
and investigate herding towards the market consensus
\citep[e.g.][]{chang2000, galariotis2015}. Our paper falls within the
latter category and tests for herding towards the market consensus for
clean energy US ETFs.

More broadly, studies that focus on herding can be broadly classified
into two main categories: on the one hand there are studies that examine
institutional investor herding and analyst herding using
transaction-level data
\citetext{\citealp{lakonishok1992}; \citealp[and][among
others]{sias2004institutional}}. On the other hand, there are studies
that explore herding towards the market consensus employing individual
returns and market returns
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}. In the
latter case, the most widely employed method is the cross-sectional
standard deviation of returns proposed by \citet{christie1995}. The
intuition behind this approach is that herding toward the market is
signaled when return dispersion is relatively low. Rational asset
pricing theory predicts that when stocks respond to market movements in
various degrees then dispersion will be linearly related to market
returns \citep[see also,][]{hwang2004market}. However, when investors
tend to mimic each other and herd towards the market the linear relation
between returns dispersion and aggregate market returns is no longer
valid. Under these conditions, fitting a non-linear regression
specification between a measure of returns dispersion and aggregate
market return appears as a more viable choice. Therefore, our approach
to measure herding activity is based on the non-linear regression model
by \citet{chang2000}.

Following this relevant literature, we compute dispersion of the
\(i^{th}\) ETF from the market return, which is known as the Cross
Sectional Absolute Deviation (\(CSAD_t\)) measure. Empirically the
\(CSAD_t\) is defined in the following manner:

@@

\pandocbounded{\includegraphics[keepaspectratio]{alt_energy_herding_files/figure-pdf/fig-rolling_window-1.pdf}}

@@

We observe several periods of herding behaviour as reflected in the
troughs in Figure~\ref{fig-rolling_window}. The most prominent cases of
herding occur between March and May of 2020 followed by several
instances of herding in the period that extends from March through April
of 2017 and the period of February-March of 2023. On the other side, we
derive significant moments of anti-herding behaviour in the clean energy
ETFs by observing the peaks in Figure~\ref{fig-rolling_window}. Cross
sectional dispersion appears to increase with respect to market-wide
returns, which is a sign of anti-herding behaviour on behalf of
investors around December of 2016 and later during September of 2022

\subsection{Climate risks and herding
behaviour}\label{climate-risks-and-herding-behaviour}

The behaviour of participants in energy markets is closely related to
the developments in the field of climate risks, carbon emissions, and
environmentally friendly policies. Rising climate risk is found to
increase green energy prices \citep{dutta2023}, and evidence from the
existing literature shows that climate policy uncertainty affects the
performance of clean energy stocks relative to dirty ones, making the
former outperform the later when the levels of climate policy
uncertainty are high \citep{bouri2022}. In particular, following the
implementation of the Paris agreement in November 2016, climate policy
uncertainty has become in the epicenter of interest across carbon and
energy markets. There are a few studies that attempt to quantify the
effects of uncertainty related to climate on the economy or financial
markets \citep[see inter alia,][]{gabriel2024, bolton2021, krueger2020}.
Interestingly, \citet{bua2024} developed two climate risk related
indexes namely transition and physical risk using a text-based approach
in order to study the effect of these risks in financial markets. In
this regard, \citet{bouri2023} study the impact of both physical and
climate risks on the returns and volatility of brown and green energy
stocks, carbon emission allowances, and green bonds, showing evidence
that transitional climate risk exerts a more significant impact than
physical climate risk.
